\textbf{A framework for effective application of machine learning to microbiome-based classification problems}

### 2020-01-28 submission/manuscript.tex
begum's edits

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  11pt,
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={A framework for effective application of machine learning to microbiome-based classification problems},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{helvet} % Helvetica font
\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
\usepackage[T1]{fontenc}
\usepackage[labelfont=bf]{caption}

@@

\newpage
\linenumbers

\hypertarget{abstract}{%
\subsection{Abstract}\label{abstract}}

@@

\hypertarget{importance}{%
\subsection{Importance}\label{importance}}

@@

\hypertarget{background}{%
\subsection{Background}\label{background}}

@@

\hypertarget{results}{%
\subsection{Results}\label{results}}

@@

\hypertarget{discussion}{%
\subsection{Discussion}\label{discussion}}

@@

We highlighted the importance of model interpretation to gain greater
biological insights into microbiota-associated diseases. In this study,
we showcased two different interpretation methods: ranking each OTU by
(i) their absolute weights in the trained models and (ii) their impact
on the predictive performance based on permutation importance.
Human-associated microbial communities have complex correlation
structures that create collinearity in the datasets. This can hinder our
ability to reliably interpret models because the feature weights of
correlated OTUs are influenced by one another (37). To capture all
important features, once we identify highly ranked OTUs, we should
review their relationships with other OTUs. These relationships will
help us generate new hypotheses about the ecology of the disease and
test them with follow-up experiments. When we used permutation
importance, we partially accounted for collinearity by grouping
correlated OTUs to determine their impact as a group. We grouped OTUs
that had a perfect correlation with each other; however, we could reduce
the correlation threshold to further investigate the relationships among
correlated features. By our approach, we identified 432 OTUs out of
6,920 that had perfect correlations with at least one other OTU. The
decision to establish correlation thresholds is left to researchers to
implement for their own analyses. Regardless of the threshold,
undestanding the correlation structures within the data is critical to
avoid misinterpreting the models. Such structures are likely to be a
particular problem with shotgun metagenomic datasets where collinearity
will be more pronounced due to many genes being correlated with one
another because they come from the same chromosome. Finally, true causal
mechanisms (e.g., role of microbiome in colorectal cancer) cannot be
explained solely by the highest performing machine learning model (38).
To identify the true underlying microbial factors of a disease, it is
crucial to follow up on any correlation analyses with further hypothesis
testing and experimentation for biological validation.

In this study, we did not consider all possible modeling approaches.
However, the principles highlighted throughout this study apply to other
ML modeling tasks with microbiome data. For example, we did not evaluate
multicategory classification methods to predict non-binary outcomes. We
could have trained models to differentiate between people with healthy
colons and those with adenomas or carcinomas (k=3 categories). We did
not perform this analysis because the clinically relevant diagnosis
grouping was between patients with healthy colons and those with SRNs.
Furthermore, as the number of classes increases, more samples are
required for each category to train an accurate model. We also did not
use regression-based analyses to predict a non-categorical outcome. We
have previously used such an approach to train random forest models to
predict fecal short-chain fatty acid concentrations based on microbiome
data (39). Our analysis was also limited to shallow learning methods and
did not explore deep learning methods such as neural networks. Deep
learning methods hold promise (12, 40, 41), but microbiome datasets
often suffer from having many features and small sample sizes, which
result in overfitting.

Our framework provides a reproducible structure for investigators
wanting to train, evaluate, and interpret their own ML models to
generate hypotheses regarding which OTUs might be biologically relevant.
However, deploying microbiome-based models to make clinical diagnoses or
predictions is a significantly more challenging and distinct undertaking
(42). For example, we currently lack standardized methods to collect
patient samples, generate sequence data, and report clinical data. We
are also challenged by the practical constraints of OTU-based
approaches. The de novo algorithms commonly in use are slow, require
considerable memory, and result in different OTU assignments as new data
are added (43). Finally, we also need independent validation cohorts to
test the performance of a diagnostic model. To realize the potential for
using ML approaches with microbiome data, it is necessary that we direct
our efforts to overcome these challenges.

@@

\hypertarget{materials-and-methods}{%
\subsection{Materials and Methods}\label{materials-and-methods}}

\textbf{Data collection and study population.} The original stool
samples described in our analysis were obtained from patients recruited
by Great Lakes-New England Early Detection Research Network (5). Stool
samples were provided by adults who were undergoing a scheduled
screening or surveillance colonoscopy. Participants were recruited from
Toronto (ON, Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor
(MI, USA). Patients' colonic health was visually assessed by colonoscopy
with bowel preparation and tissue histopathology of all resected
lesions. We assigned patients into two classes: those with healthy
colons and those with screen relevant neoplasias (SRNs). The healthy
class included patients with healthy colons or non-advanced adenomas,
whereas the SRN class included patients with advanced adenomas or
carcinomas (44). Patients with an adenoma greater than 1 cm, more than
three adenomas of any size, or an adenoma with villous histology were
classified as having advanced adenomas (44). There were 172 patients
with normal colonoscopies, 198 with adenomas, and 120 with carcinomas.
Of the 198 adenomas, 109 were identified as advanced adenomas. Together
261 patients were classified as healthy and 229 patients were classified
as having an SRN.

\textbf{16S rRNA gene sequencing data.} Stool samples provided by the
patients were used for 16S rRNA gene sequencing to measure bacterial
population abundances. The sequence data used in our analyses were
originally generated by Baxter et al.~(available through NCBI Sequence
Read Archive {[}SRP062005{]}, (5)). The OTU abundance table was
generated by Sze et al (45), who processed the 16S rRNA sequences in
mothur (v1.39.3) using the default quality filtering methods,
identifying and removing chimeric sequences using VSEARCH, and assigning
to OTUs at 97\% similarity using the OptiClust algorithm (43, 46, 47);
(\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_mBio_2018/blob/master/data/process/baxter/baxter.0.03.subsample.shared}).
These OTU abundances were the features we used to predict colorectal
health of the patients. There were 6920 OTUs. OTU abundances were
subsampled to the size of the smallest sample and normalized across
samples such that the highest abundance of each OTU would be 1, and the
lowest would be 0.

@@

For hyperparameter selection, we started with a granular grid search.
Then we narrowed and fine-tuned the range of each hyperparameter.For
L2-regularized logistic regression, L1 and L2-regularized SVM with
linear and radial basis function kernels, we tuned the cost
hyperparameter, which controls the regularization strength, where
smaller values specify stronger regularization. For SVM with radial
basis function kernel, we also tuned the sigma hyperparameter, which
determines the reach of a single training instance where, for a high
value of sigma, the SVM decision boundary will be dependent on the
points that are closest to the decision boundary. For the decision tree
model, we tuned the depth of the tree where the deeper the tree, the
more splits it has. For random forest, we tuned the number of features
to consider when looking for the best tree split. For XGBoost, we tuned
the learning rate and the fraction of samples used for fitting the
individual base learners. Performing a grid search for hyperparameter
selection might not be feasible for when there are more than two
hyperparameters to tune for. In such cases, it is more efficient to use
random search or recently developed tools such as Hyperband to identify
good hyperparameter configurations (48).

@@

\hypertarget{references}{%
\subsection{References}\label{references}}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-segata_metagenomic_2011}{}%
1. \textbf{Segata N}, \textbf{Izard J}, \textbf{Waldron L},
\textbf{Gevers D}, \textbf{Miropolsky L}, \textbf{Garrett WS},
\textbf{Huttenhower C}. 2011. Metagenomic biomarker discovery and
explanation. Genome Biol \textbf{12}:R60.
doi:\href{https://doi.org/10.1186/gb-2011-12-6-r60}{10.1186/gb-2011-12-6-r60}.

\leavevmode\hypertarget{ref-zeller_potential_2014}{}%
2. \textbf{Zeller G}, \textbf{Tap J}, \textbf{Voigt AY},
\textbf{Sunagawa S}, \textbf{Kultima JR}, \textbf{Costea PI},
\textbf{Amiot A}, \textbf{Böhm J}, \textbf{Brunetti F},
\textbf{Habermann N}, \textbf{Hercog R}, \textbf{Koch M},
\textbf{Luciani A}, \textbf{Mende DR}, \textbf{Schneider MA},
\textbf{Schrotz-King P}, \textbf{Tournigand C}, \textbf{Tran Van Nhieu
J}, \textbf{Yamada T}, \textbf{Zimmermann J}, \textbf{Benes V},
\textbf{Kloor M}, \textbf{Ulrich CM}, \textbf{Knebel Doeberitz M von},
\textbf{Sobhani I}, \textbf{Bork P}. 2014. Potential of fecal microbiota
for early-stage detection of colorectal cancer. Mol Syst Biol
\textbf{10}.
doi:\href{https://doi.org/10.15252/msb.20145645}{10.15252/msb.20145645}.

\leavevmode\hypertarget{ref-zackular_human_2014}{}%
3. \textbf{Zackular JP}, \textbf{Rogers MAM}, \textbf{Ruffin MT},
\textbf{Schloss PD}. 2014. The human gut microbiome as a screening tool
for colorectal cancer. Cancer Prev Res \textbf{7}:1112--1121.
doi:\href{https://doi.org/10.1158/1940-6207.CAPR-14-0129}{10.1158/1940-6207.CAPR-14-0129}.

\leavevmode\hypertarget{ref-baxter_dna_2016}{}%
4. \textbf{Baxter NT}, \textbf{Koumpouras CC}, \textbf{Rogers MAM},
\textbf{Ruffin MT}, \textbf{Schloss PD}. 2016. DNA from fecal
immunochemical test can replace stool for detection of colonic lesions
using a microbiota-based model. Microbiome \textbf{4}.
doi:\href{https://doi.org/10.1186/s40168-016-0205-y}{10.1186/s40168-016-0205-y}.

\leavevmode\hypertarget{ref-baxter_microbiota-based_2016}{}%
5. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
\textbf{Schloss PD}. 2016. Microbiota-based model improves the
sensitivity of fecal immunochemical test for detecting colonic lesions.
Genome Medicine \textbf{8}:37.
doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.

\leavevmode\hypertarget{ref-hale_shifts_2017}{}%
6. \textbf{Hale VL}, \textbf{Chen J}, \textbf{Johnson S},
\textbf{Harrington SC}, \textbf{Yab TC}, \textbf{Smyrk TC},
\textbf{Nelson H}, \textbf{Boardman LA}, \textbf{Druliner BR},
\textbf{Levin TR}, \textbf{Rex DK}, \textbf{Ahnen DJ}, \textbf{Lance P},
\textbf{Ahlquist DA}, \textbf{Chia N}. 2017. Shifts in the fecal
microbiota associated with adenomatous polyps. Cancer Epidemiol
Biomarkers Prev \textbf{26}:85--94.
doi:\href{https://doi.org/10.1158/1055-9965.EPI-16-0337}{10.1158/1055-9965.EPI-16-0337}.

\leavevmode\hypertarget{ref-pasolli_machine_2016}{}%
7. \textbf{Pasolli E}, \textbf{Truong DT}, \textbf{Malik F},
\textbf{Waldron L}, \textbf{Segata N}. 2016. Machine learning
meta-analysis of large metagenomic datasets: Tools and biological
insights. PLoS Comput Biol \textbf{12}.
doi:\href{https://doi.org/10.1371/journal.pcbi.1004977}{10.1371/journal.pcbi.1004977}.

\leavevmode\hypertarget{ref-sze_looking_2016}{}%
8. \textbf{Sze MA}, \textbf{Schloss PD}. 2016. Looking for a signal in
the noise: Revisiting obesity and the microbiome. mBio \textbf{7}.
doi:\href{https://doi.org/10.1128/mBio.01018-16}{10.1128/mBio.01018-16}.

\leavevmode\hypertarget{ref-walters_meta-analyses_2014}{}%
9. \textbf{Walters WA}, \textbf{Xu Z}, \textbf{Knight R}. 2014.
Meta-analyses of human gut microbes associated with obesity and IBD.
FEBS Lett \textbf{588}:4223--4233.
doi:\href{https://doi.org/10.1016/j.febslet.2014.09.039}{10.1016/j.febslet.2014.09.039}.

\leavevmode\hypertarget{ref-vazquez-baeza_guiding_2018}{}%
10. \textbf{Vázquez-Baeza Y}, \textbf{Gonzalez A}, \textbf{Xu ZZ},
\textbf{Washburne A}, \textbf{Herfarth HH}, \textbf{Sartor RB},
\textbf{Knight R}. 2018. Guiding longitudinal sampling in IBD cohorts.
Gut \textbf{67}:1743--1745.
doi:\href{https://doi.org/10.1136/gutjnl-2017-315352}{10.1136/gutjnl-2017-315352}.

\leavevmode\hypertarget{ref-qin_alterations_2014}{}%
11. \textbf{Qin N}, \textbf{Yang F}, \textbf{Li A}, \textbf{Prifti E},
\textbf{Chen Y}, \textbf{Shao L}, \textbf{Guo J}, \textbf{Le Chatelier
E}, \textbf{Yao J}, \textbf{Wu L}, \textbf{Zhou J}, \textbf{Ni S},
\textbf{Liu L}, \textbf{Pons N}, \textbf{Batto JM}, \textbf{Kennedy SP},
\textbf{Leonard P}, \textbf{Yuan C}, \textbf{Ding W}, \textbf{Chen Y},
\textbf{Hu X}, \textbf{Zheng B}, \textbf{Qian G}, \textbf{Xu W},
\textbf{Ehrlich SD}, \textbf{Zheng S}, \textbf{Li L}. 2014. Alterations
of the human gut microbiome in liver cirrhosis. Nature
\textbf{513}:59--64.
doi:\href{https://doi.org/10.1038/nature13568}{10.1038/nature13568}.

\leavevmode\hypertarget{ref-geman_deep_2018}{}%
12. \textbf{Geman O}, \textbf{Chiuchisan I}, \textbf{Covasa M},
\textbf{Doloc C}, \textbf{Milici M-R}, \textbf{Milici L-D}. 2018. Deep
learning tools for human microbiome big data, pp. 265--275. \emph{In}
Balas, VE, Jain, LC, Balas, MM (eds.), Soft computing applications.
Springer International Publishing.

>>>>

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage[unicode=true]{hyperref}
\hypersetup{
            pdftitle={A framework for effective application of machine learning to microbiome-based classification problems},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage[margin=1in]{geometry}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{helvet} % Helvetica font
\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
\usepackage[T1]{fontenc}
\usepackage[labelfont=bf]{caption}

@@

\linenumbers

\subsection{Abstract}\label{abstract}

@@

\subsection{Importance}\label{importance}

@@

\subsection{Background}\label{background}

@@

\subsection{Results}\label{results}

@@

\subsection{Discussion}\label{discussion}

@@

We highlighted the importance of model interpretation to gain greater
biological insights into microbiota-associated diseases. In this study,
we showcased two different interpretation methods: ranking each OTU by
(i) their absolute weights in the trained models and (ii) their impact
on the predictive performance based on permutation importance. Previous
studies have emphasized the difficulty of interpreting the feature
coefficients in linear models (37) and the biases introduced by
computing feature importance using built-in methods (e.g., gini drop) of
tree-based models (38). Therefore, we encourage our audience to use both
interpretation methods highlighted in this study. Human-associated
microbial communities have complex correlation structures that create
collinearity in the datasets. This can hinder our ability to reliably
interpret models because the feature weights of correlated OTUs are
influenced by one another (39). To capture all important features, once
we identify highly ranked OTUs, we should review their relationships
with other OTUs. These relationships will help us generate new
hypotheses about the ecology of the disease and test them with follow-up
experiments. When we used permutation importance, we partially accounted
for collinearity by grouping correlated OTUs to determine their impact
as a group. We grouped OTUs that had a perfect correlation with each
other; however, we could reduce the correlation threshold to further
investigate the relationships among correlated features. By our
approach, we identified 432 OTUs out of 6,920 that had perfect
correlations with at least one other OTU. The decision to establish
correlation thresholds is left to researchers to implement for their own
analyses. Regardless of the threshold, undestanding the correlation
structures within the data is critical to avoid misinterpreting the
models. Such structures are likely to be a particular problem with
shotgun metagenomic datasets where collinearity will be more pronounced
due to many genes being correlated with one another because they come
from the same chromosome. Finally, true causal mechanisms (e.g., role of
microbiome in colorectal cancer) cannot be explained solely by the
highest performing machine learning model (40). To identify the true
underlying microbial factors of a disease, it is crucial to follow up on
any correlation analyses with further hypothesis testing and
experimentation for biological validation.

In this study, we did not consider all possible modeling approaches.
However, the principles highlighted throughout this study apply to other
ML modeling tasks with microbiome data. For example, we did not evaluate
multicategory classification methods to predict non-binary outcomes. We
could have trained models to differentiate between people with healthy
colons and those with adenomas or carcinomas (k=3 categories). We did
not perform this analysis because the clinically relevant diagnosis
grouping was between patients with healthy colons and those with SRNs.
Furthermore, as the number of classes increases, more samples are
required for each category to train an accurate model. We also did not
use regression-based analyses to predict a non-categorical outcome. We
have previously used such an approach to train random forest models to
predict fecal short-chain fatty acid concentrations based on microbiome
data (41). Our analysis was also limited to shallow learning methods and
did not explore deep learning methods such as neural networks. Deep
learning methods hold promise (12, 42, 43), but microbiome datasets
often suffer from having many features and small sample sizes, which
result in overfitting.

Our framework provides a reproducible structure for investigators
wanting to train, evaluate, and interpret their own ML models to
generate hypotheses regarding which OTUs might be biologically relevant.
However, deploying microbiome-based models to make clinical diagnoses or
predictions is a significantly more challenging and distinct undertaking
(44). For example, we currently lack standardized methods to collect
patient samples, generate sequence data, and report clinical data. We
are also challenged by the practical constraints of OTU-based
approaches. The de novo algorithms commonly in use are slow, require
considerable memory, and result in different OTU assignments as new data
are added (45). Finally, we also need independent validation cohorts to
test the performance of a diagnostic model. To realize the potential for
using ML approaches with microbiome data, it is necessary that we direct
our efforts to overcome these challenges.

@@

\subsection{Materials and Methods}\label{materials-and-methods}

\textbf{Data collection and study population.} The original stool
samples described in our analysis were obtained from patients recruited
by Great Lakes-New England Early Detection Research Network (5). Stool
samples were provided by adults who were undergoing a scheduled
screening or surveillance colonoscopy. Participants were recruited from
Toronto (ON, Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor
(MI, USA). Patients' colonic health was visually assessed by colonoscopy
with bowel preparation and tissue histopathology of all resected
lesions. We assigned patients into two classes: those with healthy
colons and those with screen relevant neoplasias (SRNs). The healthy
class included patients with healthy colons or non-advanced adenomas,
whereas the SRN class included patients with advanced adenomas or
carcinomas (46). Patients with an adenoma greater than 1 cm, more than
three adenomas of any size, or an adenoma with villous histology were
classified as having advanced adenomas (46). There were 172 patients
with normal colonoscopies, 198 with adenomas, and 120 with carcinomas.
Of the 198 adenomas, 109 were identified as advanced adenomas. Together
261 patients were classified as healthy and 229 patients were classified
as having an SRN.

\textbf{16S rRNA gene sequencing data.} Stool samples provided by the
patients were used for 16S rRNA gene sequencing to measure bacterial
population abundances. The sequence data used in our analyses were
originally generated by Baxter et al. (available through NCBI Sequence
Read Archive {[}SRP062005{]}, (5)). The OTU abundance table was
generated by Sze et al (47), who processed the 16S rRNA sequences in
mothur (v1.39.3) using the default quality filtering methods,
identifying and removing chimeric sequences using VSEARCH, and assigning
to OTUs at 97\% similarity using the OptiClust algorithm (45, 48, 49);
(\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_mBio_2018/blob/master/data/process/baxter/baxter.0.03.subsample.shared}).
These OTU abundances were the features we used to predict colorectal
health of the patients. There were 6920 OTUs. OTU abundances were
subsampled to the size of the smallest sample and normalized across
samples such that the highest abundance of each OTU would be 1, and the
lowest would be 0.

@@

For hyperparameter selection, we started with a granular grid search.
Then we narrowed and fine-tuned the range of each hyperparameter.For
L2-regularized logistic regression, L1 and L2-regularized SVM with
linear and radial basis function kernels, we tuned the cost
hyperparameter, which controls the regularization strength, where
smaller values specify stronger regularization. For SVM with radial
basis function kernel, we also tuned the sigma hyperparameter, which
determines the reach of a single training instance where, for a high
value of sigma, the SVM decision boundary will be dependent on the
points that are closest to the decision boundary. For the decision tree
model, we tuned the depth of the tree where the deeper the tree, the
more splits it has. For random forest, we tuned the number of features
to consider when looking for the best tree split. For XGBoost, we tuned
the learning rate and the fraction of samples used for fitting the
individual base learners. Performing a grid search for hyperparameter
selection might not be feasible for when there are more than two
hyperparameters to tune for. In such cases, it is more efficient to use
random search or recently developed tools such as Hyperband to identify
good hyperparameter configurations (50).

@@

\subsection{References}\label{references}

\hypertarget{refs}{}
\hypertarget{ref-segata_metagenomic_2011}{}
1. \textbf{Segata N}, \textbf{Izard J}, \textbf{Waldron L},
\textbf{Gevers D}, \textbf{Miropolsky L}, \textbf{Garrett WS},
\textbf{Huttenhower C}. 2011. Metagenomic biomarker discovery and
explanation. Genome Biol \textbf{12}:R60.
doi:\href{https://doi.org/10.1186/gb-2011-12-6-r60}{10.1186/gb-2011-12-6-r60}.

\hypertarget{ref-zeller_potential_2014}{}
2. \textbf{Zeller G}, \textbf{Tap J}, \textbf{Voigt AY},
\textbf{Sunagawa S}, \textbf{Kultima JR}, \textbf{Costea PI},
\textbf{Amiot A}, \textbf{Böhm J}, \textbf{Brunetti F},
\textbf{Habermann N}, \textbf{Hercog R}, \textbf{Koch M},
\textbf{Luciani A}, \textbf{Mende DR}, \textbf{Schneider MA},
\textbf{Schrotz-King P}, \textbf{Tournigand C}, \textbf{Tran Van Nhieu
J}, \textbf{Yamada T}, \textbf{Zimmermann J}, \textbf{Benes V},
\textbf{Kloor M}, \textbf{Ulrich CM}, \textbf{Knebel Doeberitz M von},
\textbf{Sobhani I}, \textbf{Bork P}. 2014. Potential of fecal microbiota
for early-stage detection of colorectal cancer. Mol Syst Biol
\textbf{10}.
doi:\href{https://doi.org/10.15252/msb.20145645}{10.15252/msb.20145645}.

\hypertarget{ref-zackular_human_2014}{}
3. \textbf{Zackular JP}, \textbf{Rogers MAM}, \textbf{Ruffin MT},
\textbf{Schloss PD}. 2014. The human gut microbiome as a screening tool
for colorectal cancer. Cancer Prev Res \textbf{7}:1112--1121.
doi:\href{https://doi.org/10.1158/1940-6207.CAPR-14-0129}{10.1158/1940-6207.CAPR-14-0129}.

\hypertarget{ref-baxter_dna_2016}{}
4. \textbf{Baxter NT}, \textbf{Koumpouras CC}, \textbf{Rogers MAM},
\textbf{Ruffin MT}, \textbf{Schloss PD}. 2016. DNA from fecal
immunochemical test can replace stool for detection of colonic lesions
using a microbiota-based model. Microbiome \textbf{4}.
doi:\href{https://doi.org/10.1186/s40168-016-0205-y}{10.1186/s40168-016-0205-y}.

\hypertarget{ref-baxter_microbiota-based_2016}{}
5. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
\textbf{Schloss PD}. 2016. Microbiota-based model improves the
sensitivity of fecal immunochemical test for detecting colonic lesions.
Genome Medicine \textbf{8}:37.
doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.

\hypertarget{ref-hale_shifts_2017}{}
6. \textbf{Hale VL}, \textbf{Chen J}, \textbf{Johnson S},
\textbf{Harrington SC}, \textbf{Yab TC}, \textbf{Smyrk TC},
\textbf{Nelson H}, \textbf{Boardman LA}, \textbf{Druliner BR},
\textbf{Levin TR}, \textbf{Rex DK}, \textbf{Ahnen DJ}, \textbf{Lance P},
\textbf{Ahlquist DA}, \textbf{Chia N}. 2017. Shifts in the fecal
microbiota associated with adenomatous polyps. Cancer Epidemiol
Biomarkers Prev \textbf{26}:85--94.
doi:\href{https://doi.org/10.1158/1055-9965.EPI-16-0337}{10.1158/1055-9965.EPI-16-0337}.

\hypertarget{ref-pasolli_machine_2016}{}
7. \textbf{Pasolli E}, \textbf{Truong DT}, \textbf{Malik F},
\textbf{Waldron L}, \textbf{Segata N}. 2016. Machine learning
meta-analysis of large metagenomic datasets: Tools and biological
insights. PLoS Comput Biol \textbf{12}.
doi:\href{https://doi.org/10.1371/journal.pcbi.1004977}{10.1371/journal.pcbi.1004977}.

\hypertarget{ref-sze_looking_2016}{}
8. \textbf{Sze MA}, \textbf{Schloss PD}. 2016. Looking for a signal in
the noise: Revisiting obesity and the microbiome. mBio \textbf{7}.
doi:\href{https://doi.org/10.1128/mBio.01018-16}{10.1128/mBio.01018-16}.

\hypertarget{ref-walters_meta-analyses_2014}{}
9. \textbf{Walters WA}, \textbf{Xu Z}, \textbf{Knight R}. 2014.
Meta-analyses of human gut microbes associated with obesity and IBD.
FEBS Lett \textbf{588}:4223--4233.
doi:\href{https://doi.org/10.1016/j.febslet.2014.09.039}{10.1016/j.febslet.2014.09.039}.

\hypertarget{ref-vazquez-baeza_guiding_2018}{}
10. \textbf{Vázquez-Baeza Y}, \textbf{Gonzalez A}, \textbf{Xu ZZ},
\textbf{Washburne A}, \textbf{Herfarth HH}, \textbf{Sartor RB},
\textbf{Knight R}. 2018. Guiding longitudinal sampling in IBD cohorts.
Gut \textbf{67}:1743--1745.
doi:\href{https://doi.org/10.1136/gutjnl-2017-315352}{10.1136/gutjnl-2017-315352}.

\hypertarget{ref-qin_alterations_2014}{}
11. \textbf{Qin N}, \textbf{Yang F}, \textbf{Li A}, \textbf{Prifti E},
\textbf{Chen Y}, \textbf{Shao L}, \textbf{Guo J}, \textbf{Le Chatelier
E}, \textbf{Yao J}, \textbf{Wu L}, \textbf{Zhou J}, \textbf{Ni S},
\textbf{Liu L}, \textbf{Pons N}, \textbf{Batto JM}, \textbf{Kennedy SP},
\textbf{Leonard P}, \textbf{Yuan C}, \textbf{Ding W}, \textbf{Chen Y},
\textbf{Hu X}, \textbf{Zheng B}, \textbf{Qian G}, \textbf{Xu W},
\textbf{Ehrlich SD}, \textbf{Zheng S}, \textbf{Li L}. 2014. Alterations
of the human gut microbiome in liver cirrhosis. Nature
\textbf{513}:59--64.
doi:\href{https://doi.org/10.1038/nature13568}{10.1038/nature13568}.

\hypertarget{ref-geman_deep_2018}{}
12. \textbf{Geman O}, \textbf{Chiuchisan I}, \textbf{Covasa M},
\textbf{Doloc C}, \textbf{Milici M-R}, \textbf{Milici L-D}. 2018. Deep
learning tools for human microbiome big data, pp. 265--275. \emph{In}
Balas, VE, Jain, LC, Balas, MM (eds.), Soft computing applications.
Springer International Publishing.

\hypertarget{ref-thaiss_persistent_2016}{}
13. \textbf{Thaiss CA}, \textbf{Itav S}, \textbf{Rothschild D},
\textbf{Meijer MT}, \textbf{Levy M}, \textbf{Moresi C},
\textbf{Dohnalová L}, \textbf{Braverman S}, \textbf{Rozin S},
\textbf{Malitsky S}, \textbf{Dori-Bachash M}, \textbf{Kuperman Y},
\textbf{Biton I}, \textbf{Gertler A}, \textbf{Harmelin A},
\textbf{Shapiro H}, \textbf{Halpern Z}, \textbf{Aharoni A},
\textbf{Segal E}, \textbf{Elinav E}. 2016. Persistent microbiome
alterations modulate the rate of post-dieting weight regain. Nature
\textbf{540}:544--551.
doi:\href{https://doi.org/10.1038/nature20796}{10.1038/nature20796}.

\hypertarget{ref-dadkhah_gut_2019}{}
14. \textbf{Dadkhah E}, \textbf{Sikaroodi M}, \textbf{Korman L},
\textbf{Hardi R}, \textbf{Baybick J}, \textbf{Hanzel D}, \textbf{Kuehn
G}, \textbf{Kuehn T}, \textbf{Gillevet PM}. 2019. Gut microbiome
identifies risk for colorectal polyps. BMJ Open Gastroenterology
\textbf{6}:e000297.
doi:\href{https://doi.org/10.1136/bmjgast-2019-000297}{10.1136/bmjgast-2019-000297}.
